GEE models

### 2017-10-02 doc/submission/reviewer-response-2.md

# Response to reviewers

## Referee: 1

- *"This paper presents an investigation of the relationship between NEFAs measured
at baseline and subsequent development of markers of insulin sensitivity. The
analysis is very thorough and the conclusions, in particular the interpretation
of results with biological considerations is excellent. I am not an expert on
this subject area and although I am a statistician, GEE is not an approach that
I am familiar with, so my comments are from the perspective of a
technically-minded but non-expert reader. Therefore some of my comments may not
be wholly relevant and are given more as possible considerations; or areas that
may require clarification for other non-experts."*

    We thank the reviewer for the comments and hope that the below responses
    adequately address issues or suggestions raised.

- *"A key question for me is the choice of using the GEE approach, which is of
course an established and appropriate method for many research questions, but I
think could be better justified here. The main advantages I am aware of are:
Richer correlation structures, easily handles missing data, robust standard
errors. Here, the number of follow up visits is not large and the amount of
missing data is relatively small. My first thought when approaching the analysis
would therefore be to consider a standard linear regression analysis with
robust/clustered standard errors to incorporate repeated measures, and potential
multiple imputation of missing values and a comparison with complete case
analysis. So it would be nice to briefly spell out the things that GEE offers
here over other approaches to make this clear (in the ESM if need be so as not
to add to the word count). Currently the main statement is the ease of handling
missing data - which is handy but perhaps not essential - and some comments on
the working correlation matrix, which are not terribly informative. So some
statements about the improvement in estimates under a specific correlation
structure that is not easily handled otherwise, for instance, might be helpful.
Of note is that in the ESM it is stated that "other ... matrices had similar fit"
- it may be worth noting that the model fit may be similar, but estimates may be
more efficient under a particular correlation structure."*

    The reviewer's suggestion to more fully describe the justification for using
    GEE over other methods is warranted and we will expand on this in the
    manuscript in the Methods and the ESM Methods sections. We chose the GEE
    technique for a variety of reasons. As the reviewer states, GEE is able to
    handle visit dates that differ between participants (e.g. some participants
    come exactly 3 years later for a followup while others come some months
    before or after 3 years) and for dealing with the inherent correlational
    structure underlying longitudinal data. Because of these features, among
    others, GEE can compute robust standard errors of the estimates that more
    accurate reflect the underlying data. The ability to specify a working
    correlation matrix makes this a powerful technique for longitudinal data
    compared to standard linear regression, as it removes the underlying
    assumption that variables are independent (which is not the case with
    longitudinal data). While choosing the specific correlation matrix is
    important, GEE still produces robust estimates and standard errors even with
    a misspecified correlation matrix. In the case of longitudinal data, an
    auto-regressive correlation matrix is the most appropriate statistically and
    biologically since measurements on an individual over time will change (e.g.
    people gain or lose weight over time).

- *"Another thing that might be commented on briefly is the different nature of
estimates from GEE models, which are population averaged and have a slightly
different interpretation. This difference might be seen as a disadvantage to GEE
modelling - or do the authors feel this is more appropriate for this particular
question? This is something I do not have a thorough understanding of, but I do
note that the authors comment in the methods "...the interpretation of the GEE
results is as an expected percent difference in the outcome variable for every
standard deviation (SD) increase in the predictor variable given the covariates
are held constant" does not appear to be consistent with what I have read in
terms of keeping other factors constant, e.g.: "However, because the GEE
approach does not contain explicit terms for the between-cluster variation, the
resulting parameter estimates for the contrast of interest do not have the usual
"keeping other factors constant" interpretation." (Hanley et al, Am J Epidemiol
2003;157:364–375)."*

    We were only concerned with population level averages, and not
    individual-level estimates. GEE is a marginal model, which estimates the
    population average, and is one of the other reasons we chose GEE. However,
    even though GEE is a marginal model, for continuous outcomes (e.g. insulin
    sensitivity) the estimates derived from GEE's counter part, mixed effects
    models, are more or less equivalent in both the value and in the
    interpretation. For nonlinear outcomes (e.g. binary values such as diabetes
    incidence), the interpretation and estimates are substantially different
    from mixed effects models. We did not model nonlinear outcomes and as such
    this difference is {{irrelevant?}}.

    Regarding the cited paper, the authors state the paper is an
    *"...orientation focused on correlated data arising from the relatedness of
    **several individuals** in the **same** cluster, rather than several
    "longitudinal" observations in the same individual"* (emphasis added). This
    is an important distinction and their statement about the "keeping other
    factors constant" and the lack of terms for the "between-cluster variation"
    is true in this case. However, this present analysis has the repeated
    measures at the individual level; the cluster is the individual and the
    within-cluster variation is the individual over time. In this context,
    interpreting the estimates derived from GEE is comparable to the classical
    linear regression equations:

    y = B~0~ + B~1~ X~1~ + ... + B~n~ X~n~

    Such that the interpretation of e.g. total NEFA with ISSI-2 is that for
    every one SD increase in total NEFA there is an average 5% lower ISSI-2 for
    individuals at the same time point and of the same age, waist, ethnicity,
    etc.

@@

- *"Another area that is unclear is the exact model specification. The unadjusted
results I can assume are the outcome variables regressed via GEE against each
NEFA measure. Then adjusted results include age, gender, WC and so on. Now, for
the proportion of NEFA results, the approach seems to be that adjusted models
were fitted with each NEFA as a predictor in turn while adjusting for the
anthropomorphic and other covariates, as a PLS approach was used to examine all
the NEFA components together as proportions. Was this approach the same for the
analysis of NEFA as a concentration, or in this analysis were all NEFA
components included in a single multivariable model? I think that this should be
spelled out: if each NEFA component was considered in turn, and the variables
adjusted for, for the concentration/proportion analysis. At present the
assumption might be that multivariable models include all of the NEFA components
in one model."*

    The reviewer brings up excellent points. The reason NEFA were included as a
    proportion in the PLS analysis rather than as a concentration was due to the
    inherent interdependence between NEFA values as a proportion. They all must
    equal to 100. It is this dependency structure in the data that allows PLS to
    more accurately identify clusters within the data. Concentration data does
    not have this inherent dependency (all NEFA values do *not* equal to 100).
    If one NEFA goes up in concentration, other NEFA values do not necessarily
    go down, as would be the case with the proportion. We have clarified this in
    the methods. {{not all of this comment is clear to me}}.

@@

- *"If so, I do not fully understand why there is not a model that includes the
joint effect of all NEFA components together. Could this not be done for the
concentration analysis? Also, it could perhaps be done for the proportion too,
if one of the components is set as the baseline – similar to how categorical
variables are handled in regression analyses, but allowing percentages for each
indicator rather than the usual 0/1 dichotomy, vs. a "baseline" component. Such
an analysis could also adjust for total NEFA, with the interpretation that
proportion of each NEFA component is assessed while accounting for other NEFA
components, and also the total NEFA. This is also something I struggled to
understand a little in general – why should *proportions* of NEFA components be
important? There may be a relatively high proportion of a particular NEFA but if
the patient has a generally low overall NEFA, is there any reason to think a
relatively high proportion of something will increase insulin sensitivity? These
ideas may be wholly implausible of course and I may have missed the point – as I
said my background knowledge of the biology is not ideal."*

    {{Don't fully understand this comment, but I'll try to answer}}. We hope the
    response to the point above partially addresses this comment. To reiterate,
    the reason NEFA proportion was used in the PLS analysis was due to the
    inherent interdependency of the values as a proportion. When they are
    proportions, PLS would be more likely to pick up potential clusters since
    changes in one value *must* lead to changes in another value as they all
    equal 100.

    The reason we did not include all NEFA variables (as proportion or as
    concentration) into a single GEE model is because GEE can not handle this
    level of complexity, or at least, the interpretation for us and the readers
    would be nearly totally meaningless or incomprehensible. How the reviewer
    described setting the variables in the regression equation may be useful for
    some research questions and with fewer variables. However, we had 22
    different fatty acids (44 if including mol% and nmol/ml) and 4 different
    outcomes. The high dimensionality of this analysis makes interpretation
    using standard linear regression approaches extremely challenging, not to
    mention introducing potential bias due to correlation among the NEFA
    (especially as a proportion). These reasons are why we chose to run multiple
    GEE models for each NEFA-outcome pairing in addition to running a PLS
    analysis with all NEFA in a single model (since PLS is designed for this
    particular analysis).

    The reviewer is correct that proportions *alone* are not enough to draw a
    meaningful conclusion, as per how the reviewer described it. This, we
    believe, is a major strength of our study since previous literature using other
    fatty acid fractions (e.g. phospholipid) generally only investigate
    proportions but do not report on or measure the concentration. Concentration
    data is more technically challenging to quantify and so is understandable
    why large cohort studies do not measure this data. Using both concentration
    and proportion, we can identify either it is *both* the total size of any
    individual NEFA and a specific proportion of a single or multiple individual
    NEFA that associates with risk for diabetes. Or whether it is only the total
    concentration, as we found in this study.

- *"I do also wonder somewhat whether there may be over-adjustment with some of the
covariates included, and agree with another reviewer who pointed out the
importance of the unadjusted results. A potential issue is confounders that may
be on the causal pathway, which is a judgement that must be based on expert
knowledge. The authors note for instance that adjusting for WC improved the fit
of the model, and was indicated for inclusion by the DAG. Neither in my mind is
reason alone for its inclusion, and there is no guarantee on this basis that
adjustment for this is necessarily appropriate - this must be based on
subject-specific knowledge of whether adjustment is appropriate to determine the
independent effect of the variable of interest."*

    {{I'm not entirely clear how to address this. Since WC hypothetically could
    lead to higher NEFA as well as lower IS (in particular), and since we are
    testing whether higher NEFA lowers IS, how can we not adjust for it? It's a
    classic confounder. I'll take a stab at answering it though.}} There may be
    some potential over-adjustment in our models. However, we have taken
    extensive steps to understand the biological and causal pathways underlying
    our associations and to empirically determine which covariates to include in
    the model.

    In the case of WC, we feel there is substantial grounds for adjusting for it
    when studying NEFA on metabolic capacity: 1) biologically NEFA comes from
    adipose tissue; 2) intra-abdominal fat tissue is more metabolically active;
    3) adipose tissue is an active endocrine organ that contributes to insulin
    sensitivity and metabolism in general; and 4) WC is a direct measure of
    abdominal (and indirectly intra-abdominal) fat mass. It is for these reasons
    that adjusting for WC is, we believe, strongly warranted.

    Having said that, because of comments from previous reviewers about this
    very issue, we included an analyses to address this issue. We ran models
    adjusted for all covariates except for WC and reported on the results in the
    ESM material. We had also included a discussion about this issue in
    conclusions, which we hope will inform the interested reader about this
    particular issue and allow them to draw their own interpretations of the
    results based on this raised issue.

@@

- *"Which brings me to the DAGs. "Directed acyclic graphs" just means a collection
of boxes representing different bits of information with arrows between them
that do not form loops; DAGs are used in Bayesian analysis for instance to
represent how data and parameters in a model are linked. A DAG therefore is a
representation for how information in linked, not a process to in itself to
determine appropriate model building or make a "recommendation". I think that if
the approach employed is to be used as justification of the model building
process, it needs to be better described: presumably there is some algorithm
being used that determines interrelationships of variables in the data, of which
I would like to see at least a brief outline – a link to a website is not really
sufficient here. Further, the graphs provided give no indication of what the
different arrows and colours are."*

    We agree with the reviewer that the wording around DAG use in model building
    was not entirely clear and we have worked to make it clearer in the ESM and
    methods. DAGs *are* in fact a process to identify potential covariates to
    include in a model when used algorithmically. This mathematically construct
    underlying DAGs and model building is the reason the DAGitty software was
    developed, so that this mathematical process could be automated. Much
    greater detail on how DAGs are used to build models can be seen in the two
    references listed below. The DAG figures in the manuscript will be described
    in more detail, including what each colour and arrow indicates.

    - Greenland S, Pearl J, Robins JM (1999) Causal diagrams for epidemiologic
    research. Epidemiology 10:37–48.
    - Textor J, Hardt J, Knüppel S (2011) DAGitty: A graphical tool for
    analyzing causal diagrams. Epidemiology 22:745.

@@

- *"Last of all, the analysis excluded diabetes cases, which is justified. However,
could incident cases be included with follow up time included to the last visit
prior to developing diabetes? This may not be so straightforward in practice of
course. It’s not really clear if any of any pre-incident follow up time is
included at present, or if all data on incident cases is discarded "incident
diabetes cases were excluded from further GEE analysis" – does the "further" bit
mean post-incidence here?"*

    Yes, all follow-up data is included *until* the visit with the incident
    diabetes case (e.g. a participant is diagnosed with diabetes at the 6 year
    visit, but the baseline and 3rd year visit data are still included in the
    analysis). This reason, among others, is why GEE and similar techniques are
    so powerful. The wording has been fixed to make this point clearer.

- *"MINOR COMMENTS:"*
    - *"Abstract and elsewhere: OGTT not spelled out, even if this is obvious to
    most readers."*

        This suggestion has been added to the abstract.

    - *"List of abbreviations: if this is to be an exhaustive list many are not
    given, OGTT, HOMA, NEFA"*

        These have been added as suggested.

@@

    - *"QIC - presumably this is quite like the AIC and other information criteria -
    maybe mention this, and how the calculation differs? Is the QIC specifically for
    GEE models? Also in the ESM: "This identifies a specific model from a number of
    models that is the 'best' fit and least complex of all the models" - it can’t be
    the best fit AND least complex, as mentioned in the previous sentence it’s a
    trade-off - so maybe omit this sentence."*

        The reviewer is correct that QIC is similar to AIC. We have expanded on
        the explanation about QIC vs other information criterion. The reviewer
        is correct that QIC is specific to GEE-type models. Because GEE uses
        quasi-likelihood rather than maximum likelihood estimation (as other
        techniques use such as mixed effects models) information criteria such
        as AIC can not be used since they rely on maximum likelihood. We agree
        that the wording surrounding 'fit' and 'complex' could be clearer.
        However, we disagree that it should be omitted since techniques such as
        QIC *do* select models that have both a good fit and are less complex.
        While techniques such as AIC are powerful model selection methods, they
        have an inherent limitation in that models with more parameters will
        have a better fit. So models with more parameters (more 'complex') be
        more likely to be selected by default when using AIC. QIC and other
        similar information criteria techniques on the other hand try to balance
        this limitation by explicitly including terms that penalize a more
        'complex' model (more parameters) while still emphasizing a good fit. We
        have clarified this in the manuscript with the wording: "... that has
        both a good model fit while keeping model complexity low (less variables
        included)."

@@

        We have added this to the ESM Methods section.

## Referee: 2

- *"In this manuscript, the authors investigated the link between total and several
fractions of serum NEFA and the clinical indexes of insulin secretion and
sensitivity by using GEE model in a diabetes risk group over 6-years
longitudinal follow-up."* ... *"Overall, this study is interesting and novel to
show the long-term exposures to some specific NEFA fractions are negatively
associated with mainly beta-cell function in human, which convincingly confirms
and advances the previous basic and clinical studies in this field."*

    We thank the reviewer for this comments and the suggested edits below. We
    have incorporated most of the suggestions and have provided responses that
    hopefully clarify other questions and comments.

@@

1. *"Serum insulin and/or NEFA levels could be affected by several factors: I
would suggest excluding the participants who have endocrine, hepatic or renal
disorders. Also, I suggest mentioning the detail of medications, which could
affect insulin/glucose/NEFA measurement; page 24, line 7, "Medication use did
not influence NEFA composition, the outcomes, or the GEE models (data not
shown)" is not clear enough."*

    {{confirm that participants didn't have other disease}}. We have expanded on
    the details about medication use (i.e. general types taken).

@@

2. *"Glucose clamp test is the standard to evaluate insulin sensitivity instead
of the indexes used in this study. I know it is very hard to perform it in a
large number of participants, but it may be one of the reasons why the effect of
NEFAs on insulin sensitivity was limited. I would suggest discussing the
limitation of these indexes."*

    We agree, this suggestion is pertinent and has been included in the
    limitations section of the conclusion.

@@

3. *"Most of the indexes in this study are based on glucose and insulin levels.
I think the reader would know the data of baseline and follow-up glucose and
insulin levels as well as indexes (in Table 1)."*

    Both glucose and insulin have been added to Table 1.

4. *"I understand several NEFA fractions are statistically related to indexes of
beta-cell function; however, does each individual show the same trend? For
example, the graph which shows NEFA concentrations and the individual decline
ratio of beta-cell function over 6-years (value of slope, if compatible) may
convince the reader more, at least for the representative NEFA fractions."*

    {{This comment is not entirely clear to me}} This is an interesting
    question. GEE does not model on the individual level, only on the population
    as a whole. As such, we cannot answer research questions specific to
    individuals. There are statistical techniques that can extract individual
    level associations, however, our objective was (at least initially) to
    examine population level associations. We therefore chose statistical
    techniques suitable to the study design and our research questions, which
    was in this case GEE modelling.

@@

5. *"Referring previous longitudinal studies showing the link between NEFA level
and beta cell deterioration, even which were investigated in the diabetic
patients, could confirm the results of this study."*

    We have added some relevant references and associated discussions to the
    conclusions.

@@

## Referee: 3

- *"The authors addressed all major concerns related to the manuscript. I feel
that the paper improved by the changes and became more focused by omitting the
PLS-DA analysis. By adding a detailed explanation to the discussion on the
potential importance of potential confounders and mediators, the interpretation
of the findings became much clearer."*

    We thank the reviewer for the comments and have made the suggested changes
    as requested below.

@@

## Referee: 4

- *"This is an interesting analysis on NEFA ,insulin sensitivity and beta cell
function."* ... *"My major concern is that with the current approach the
complexity of metabolic regulation may be lost to some extent. This makes it
hard to interpret these findings with respect to biological significance:"*

    - *"With respect to peripheral insulin sensitivity (referred to in the
    discussion) the effects of saturated fatty acids may strongly depend on
    metabolic phenotype, ie more pronounced in insulin resistant subjects, was
    initial metabolic phenotype taken into account in the present analysis?"*

    The initial metabolic phenotype and subsequent influence of saturated fatty
    acids is taken into account in the time-by-fatty acid interaction. {{expand}}

@@

    - *"it is a pity that the analysis of fatty acids composition in
    phospholipids and serum cholesteryl esters have been published in a separate
    paper, altogether this would add very much to the total interpretation."*

### 2017-10-02 doc/submission/reviewer-response-2.md

    While we agree with the reviewer that inclusion of the phospholipid and
    cholesteryl ester farctions would have added to the interpretation, the
    volume and complexity of the analyses would not have been publishable with
    current journal word and figure count limits. This was the reason we decided
    to publish the fractions in separate manuscripts, to maintain a clear and
    focus narrative within the constraints of the journals limits {{rephrase}}.
    {{mention my thesis?}}

@@

- *"Altogether, this paper would gain strength by addressing the limitations in
interpretation in more detail."*

## Associate Editor

- *"The reviewers provide numerous minor and major comments, for example regarding
motivation of the statistical approach, which require revision or careful
response."*

    We hope that the revisions made and the responses to the reviewers
    adequately address all of their concerns or comments.

### 2017-10-02 doc/manuscript.Rmd
superscript affil
remove fax number for tony

<!-- First, middle initial, last -->
Author: Luke W. Johnston (1); Stewart B. Harris (2); Ravi Retnakaran
(3,4); Adria Giacca (5); Zhen Liu (1);
Richard P. Bazinet (1); and Anthony J. Hanley (1,3,6)

@@

- Name: Anthony J. Hanley
- Current address:  
    Department of Nutritional Sciences  
    Faculty of Medicine  
    University of Toronto  
    FitzGerald Building, 150 College Street, Room 341  
    Toronto, ON, Canada, M5S 3E2
- Phone number: 416.978.3616
- Fax number:
- Email: anthony.hanley@utoronto.ca

>>>>

<!-- First, middle initial, last -->
Author: Luke W. Johnston^1^; Stewart B. Harris^2^; Ravi Retnakaran
^3,4^; Adria Giacca^5^; Zhen Liu^1^;
Richard P. Bazinet^1^; and Anthony J. Hanley^1,3,6^

@@

- Name: Anthony J. Hanley
- Current address:  
    Department of Nutritional Sciences  
    Faculty of Medicine  
    University of Toronto  
    FitzGerald Building, 150 College Street, Room 341  
    Toronto, ON, Canada, M5S 3E2
- Phone number: 416.978.3616
- Email: anthony.hanley@utoronto.ca
